Typed Revision Control

### 2024-09-12 mss.tex
Post

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              Manuscript submitted for review.\\
              % {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Typed Revision Control}
\fancyfoot[LE,RO]{\thepage}

@@

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{1}

@@

\section[{A History of Type in {\sc rcs}s}]{A History of Type in RCSs}

@@

\noindent
The employment of an intermediate representation allows us to track changes to files and manage their history—state—more effectively.  Consider the popular JavaScript framework React \citep{React}, aspects of which were developed to resolve difficulties with slow repainting in browsers (for a popular explanation, see \citeauthor{Baer2018} (\citeyear{Baer2018}), chapter 1).  React's virtual Document Object Model ({\sc dom}) system acts as an intermediate representation between the actual {\sc dom} and the developer's code.  When a developer updates the virtual {\sc dom}, the framework compares the new virtual {\sc dom} with the previous virtual {\sc dom} and updates the real {\sc dom} only if necessary.  This allows React to optimize the rendering process and improve performance.  It also means that we can consider the structural representation of the {\sc dom} object as an intermediate representation, and the production and application of the diff as a type-aware {\sc rcs} operation.  Higher-level interfaces like Microsoft Word's “Track Changes” feature \citep{MS2024} similarly resolves diffs across text and formatting within a document, an artifact processed from a file into a user-friendly representation.  While the “Track Changes” user interface is optimized for interactivity, conceptually it is quite similar to conventional merge tools from {\sc rcs}s, and de facto resolves the same sorts of diffs for its particular file type.  (Indeed, one can imagine a Word-aware command-line {\sc rcs} that could meaningfully merge changes to a Word document, rather than simply storing successive copies of the file.)  Thus a significant part of managing typed file data is deciding what the input, output, and diffs should look like, and the remainder is systematic application of these rules to the data.

@@

\noindent
Strongly-typed version control requires coupling to the build system, since the system must be aware of the structure of the data in order to apply diffs.  (Indeed, \citeauthor{Blackman2024} (\citeyear{Blackman2024}), pp. 6–7, in {\sc ustj} vol. 1 iss. 1, pp. 1—46, discussed one solution to the linking problem which is well-suited to Urbit's global namespace.)

@@

\noindent
Marks do not need to be symmetrical or round-trip.  A {\sc json} input may yield an equivalent {\sc json} output but altered by structural whitespace.  Clay is also proactive in searching for possible mark paths—if an immediate conversion is not available, then a mutually available intermediary form may be used.  To implement a custom filetype, all that is required is the production of a suitable mark file.  Conversions from the filesystem require at a minimum a {\sc mime} mark file, which acts as the intermediary IR from the host OS.

@@

A desk is a filesystem continuity,\footnote{See \citet{Blackman2024a} in {\sc ustj} vol. 1 iss. 1, pp. 75—82, section 2.2 for an exposition of the so-called ``theory of a desk''.} and represents the main way that data are synchronized between Urbit instances (ships).  The state of a desk is a history of sequential commits, with each commit after the first having at least one parent commit.  Desks are often analogized to Git repositories.\footnote{Although desks are synchronized to the host OS as folders, this is a merely a convention, and other schemes of noun organization could employ other organizational abstractions.}  Each desk must contain sufficient internal data—especially marks—to produce its own code, given Hoon and the standard library (which are always available).

@@

In principle, as we have intimated, an {\sc rcs} may abstract away from the file as the fundamental artifact being tracked.  While Clay does not fully commit to this, in a concession to its {\sc posix}-compliant host OS, it gestures towards the possibility of conceiving of build sources and artifacts as objects (nouns) other than files.  Since Urbit ``unifies'' nouns (that is, it structurally stores only one copy of a noun and maintains references to that shared value), repetition of critical resources across desks does not lead to filesystem bloat.  (See \citet{Blackman2024a} in {\sc ustj} vol. 1 iss. 1, pp. 75—82, for details.)

@@

In its current instantiation, Clay exhibits some notable characteristics as a typed revision control system.  It is capable of tracking file changes at a structural level, permitting meaningful diffs and merges for files of the same mark.  It is also capable of converting between marks, permitting a more flexible approach to file type than the conventional text/binary dichotomy of Git and other {\sc rcs}s.  However, the future of userspace and code building is still in flux and may see a reworking of the Clay vane to better suit the needs of developers and users.

>>>>

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:2 (2024):  33–48. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Typed Revision Control}
\fancyfoot[LE,RO]{\thepage}

@@

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{33}

@@

\section{Type in Revision Control Systems}

@@

\noindent
The employment of an intermediate representation allows us to track changes to files and manage their history—state—more effectively.  Consider the popular JavaScript framework React \citep{React}, aspects of which were developed to resolve difficulties with slow repainting in browsers (for a popular explanation, see \citeauthor{Baer2018} (\citeyear{Baer2018}), chapter 1).  React's virtual Document Object Model ({\sc dom}) system acts as an intermediate representation between the actual {\sc dom} and the developer's code.  When a developer updates the virtual {\sc dom}, the framework compares the new virtual {\sc dom} with the previous virtual {\sc dom} and updates the real {\sc dom} only if necessary.  This allows React to optimize the rendering process and improve performance.  It also means that we can consider the structural representation of the {\sc dom} object as an intermediate representation, and the production and application of the diff as a type-aware {\sc rcs} operation.  Higher-level interfaces like Microsoft Word's “Track Changes” feature \citep{MS2024} similarly resolves diffs across text and formatting within a document, an artifact processed from a file into a user-friendly representation.  While the “Track Changes” user interface is optimized for interactivity, conceptually it is quite similar to conventional merge tools from {\sc rcs}s, and de facto resolves the same sorts of diffs for its particular file type.\footnote{Indeed, one can imagine a Word-aware command-line {\sc rcs} that could meaningfully merge changes to a Word document, rather than simply storing successive copies of the file.  The OpenDocument file form (\texttt{odt}, \texttt{ods}, \texttt{odg}, etc.) consists of \textsc{ZIP}-compressed \textsc{xml} files, which would be highly amenable to type-aware merge tools at the command line.}  Furthermore, the ``Source Code in Database'' (\textsc{scid}) approach parses all code into a database as an intermediate representation; this shares some features with a hypothetical type-aware \textsc{rcs}.  Thus a significant part of managing typed file data is deciding what the input, output, and diffs should look like, and the remainder is systematic application of these rules to the data.

@@

\noindent
Strongly-typed version control requires coupling to the build system, since the system must be aware of the structure of the data in order to apply diffs.  (Indeed, \citeauthor{Blackman2024} (\citeyear{Blackman2024}), pp. 6–7, in \ssc{ustj} vol. 1 iss. 1, pp. 1—46, discussed one solution to the linking problem which is well-suited to Urbit's global namespace.)

@@

\noindent
Marks do not need to be symmetrical or round-trip.  A {\sc json} input may yield an equivalent {\sc json} output but altered by structural whitespace.  Clay is also proactive in searching for possible mark paths—if an immediate conversion is not available, then a mutually available intermediary form may be used.  To implement a custom filetype, all that is required is the production of a suitable mark file.  Conversions from the filesystem require at a minimum a {\sc mime} mark file, which acts as the global intermediary from the host OS.

@@

A desk is a filesystem continuity,\footnote{See \citet{Blackman2024} in \ssc{ustj} vol. 1 iss. 1, pp. 1—46, section 2.2 for an exposition of the so-called ``theory of a desk''.} and represents the main way that data are synchronized between Urbit instances (ships).  The state of a desk is a history of sequential commits, with each commit after the first having at least one parent commit.  Desks are often analogized to Git repositories.\footnote{Although desks are synchronized to the host OS as folders, this is a merely a convention, and other schemes of noun organization could employ other organizational abstractions.}  Each desk must contain sufficient internal data—especially marks—to produce its own code, given Hoon and the standard library (which are always available).

@@

In principle, as we have intimated, an {\sc rcs} may abstract away from the file as the fundamental artifact being tracked.  While Clay does not fully commit to this, in a concession to its {\sc posix}-compliant host OS, it gestures towards the possibility of conceiving of build sources and artifacts as objects (nouns) other than files.  Since Urbit ``unifies'' nouns (that is, it structurally stores only one copy of a noun and maintains references to that shared value), repetition of critical resources across desks does not lead to filesystem bloat.  (See \citet{Blackman2024a} in \ssc{ustj} vol. 1 iss. 1, pp. 75—82, for details.)

@@

In its current instantiation, Clay exhibits some notable characteristics as a typed revision control system.  It is capable of tracking file changes at a structural level, permitting meaningful diffs and merges for files of the same mark.  It is also capable of converting between marks, permitting a more flexible approach to file type than the conventional text/binary dichotomy of Git and other {\sc rcs}s.  However, the future of userspace and code building is still in flux and may see a reworking of the Clay vane to better suit the needs of developers and users.\tombstone{}
